$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-09-07 paper/paper.tex
Move the example process to the introduction

@@

\autoref{fig:algebra-instances} sketches the implementation of linear, graded and shared types as instances of our usage algebra.
Their use in typing derivations is illustrated in \autoref{example-derivations}.

@@

\begin{example}[Typing derivation]
\label{example-derivations}

@@

\begin{flalign*}
& \func{sync} \; : \; \Var_n && \to \Var_n \to \Var_n \to \Process_n & \\
& \func{sync} \; in0 \; in1 \; out \; &&= \\
& && \recv{in0}{} \\
& && \recv{(\suc \; in1)}{} \\
& && \send{(\suc \suc \; out)}{\suc \constr{0}} \\
& && \send{(\suc \suc \; out)}{\constr{0}} \PO \\
\end{flalign*}
\begin{flalign*}
& \func{\vdash sync} && : \types{\gamma \comma \channel{t}{\lz} \comma \channel{t}{\lz} \comma \channel{t}{\lz}\\ & &&}{\Gamma \comma (1 \comma 0) \comma (1 \comma 0) \comma (0 \comma 2)\\ & &&}{\func{sync} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}\\ & &&}{\Gamma \comma (0 \comma 0) \comma (0 \comma 0) \comma (0 \comma 0)} & \\
& \func{\vdash sync} && = \\
& && \recv{\suc \suc \constr{0}}{} \\
& && \recv{\suc \suc \constr{0}}{} \\
& && \send{\suc \suc \constr{0}}{\suc \constr{0}} \\
& && \send{\suc \suc \constr{0}}{\constr{0}} \PO \\
\end{flalign*}

@@

\begin{flalign*}
& \func{recv} \; : \; \Var_n \to \Process_n & \\
& \func{recv} \; c \; = \recv{c}{} (\recv{\suc c} \PO) \\
& \\
& \func{\vdash recv} \; \types{\gamma \comma \channel{t}{\lz}}{\Gamma \comma (2 , 2)}{\func{recv} \; \constr{0}}{\Gamma \comma (0 \comma 2)} & \\
& \func{\vdash recv} \; = \; \recv{\constr{0}}{} \recv{(\suc \constr{0})}{} \PO \\
& \\
& \func{send} \; : \; \Var_n \to \Var_n \to \Process_n & \\
& \func{send} \; c \; v \; = \send{c}{v} \PO \\
& \\
& \func{\vdash send} \; \types{\gamma \comma \channel{t}{\lz} \comma t}{\Gamma \comma (1 , 1) \comma \lz}{\func{send} \; (\suc \constr{0}) \; \constr{0}}{\Gamma \comma (1 \comma 0)\comma \lz} & \\
& \func{\vdash send} \; = \; \send{(\suc \constr{0})}{\constr{0}} \PO
\end{flalign*}

@@

\begin{flalign*}
& \func{example} && : \; \Process_{\constr{0}} & \\
& \func{example} && = \new{} ( ( \new{} (\func{send} \; \suc \constr{0} \; \constr{0})) \\
& && \comp{}{ \new{} ( (\new{} (\func{send} \; \suc \constr{0} \; \constr{0} })) \\
& && \comp{}{ \new{} ( \func{recv} \; \constr{0} } \\
& && \comp{}{ \func{sync} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}))) } \\
\end{flalign*}
\begin{flalign*}
& \func{\vdash example} && : \; \types{\constr{[]}}{\constr{[]}}{\func{example}}{\constr{[]}} & \\
& \func{\vdash example} && = \new{} \; \_ \; \_ \; \_ \; (( \new{} \; \_ \; \_ \; \_ \; \func{\vdash send}) \\
& && \comp{}{ \new{} \; \_ \; \_ \; \_ } \; (( \new{} \; \_ \; \_ \; \_ \; \func{\vdash send}) \\
& && \comp{}{ \new{} \; \channel{\unit}{\func{\omega}} \; \_ \; \_ } \; ( \func{\vdash recv} \\
& && \comp{}{ \func{\vdash sync} }))) \\
\end{flalign*}

>>>>

\begin{example}[Example process]
\label{example-process}

We present an example that highlights the ways in which graded types can be used to prevent mistakes in the definition of \picalc{} processes.
The syntax uses Bruijn indices as variable references and is introduced in detail in \autoref{syntax}.
In short: $\new{} P$ creates a new channel and makes it available at index $0$ in $P$; $\comp{P}{Q}$ composes $P$ and $Q$ in parallel; $\recv{x}{}P$ receives along channel $x$ and makes the data available at index $0$ in $P$; and $\send{x}{y}{P}$ sends $y$ over channel $x$ and continues with $P$.
Both variable references and processes are indexed by the number of free variables they can refer to (for any natural number $n$, $\Var_n$ and $\Process_n$, respectively).
For the sake of clarity, we define the example subprocesses separately and parametrise them by the channels on which they operate.

\begin{figure}[h]
  \tiny
  \centering
  \begin{tikzpicture}[draw,circle]
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=orange] (send-x) at (0,2)   {$\func{send} \, x$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=orange] (send-y) at (2,2)   {$\func{send} \, y$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=green] (recv-z) at (4,2)   {$\func{recv} \, z$};
    \node[draw,circle,fill opacity=0.3,text opacity=1,fill=brown,align=center] (sync) at (2,0)   {$\func{sync}$\\$ \, x \, y \, z$};

    \draw[->] (send-x) -- node[left] {$x$} (sync);
    \draw[->] (send-y) -- node[right] {$y$} (sync);
    \draw[->] (sync.35) -- node[right] {$z$} (recv-z.240);
    \draw[->] (sync) -- (recv-z);
  \end{tikzpicture}
  \caption{Four communicating processes composed in parallel.}
  \label{fig:example-process}
\end{figure}

The example contains four processes that are composed in parallel: two of them send data over a channel; one receives data twice; the fourth sticks together and synchronizes the previous processes (\autoref{fig:example-process}).
The process doing the synchronisation first receives on both input channels and then sends the received data along the output channel.
Thanks for linear and graded types, the processes below have their interactions safely restricted: the usage annotation $(0 \comma 1)$ on channel $c$ (no input, output once) restricts what both $\func{send}$ subprocesses can do; and $(2 \comma 0)$ (input twice, no output) restricts what $\func{send}$ can do.
Dually, the remaining annotations $(1 \comma 0)$, $(1 \comma 0)$ and $(0 \comma 2)$ are given to the subprocess $\func{sync}$.
We use these types to give a suitable typing derivation in \autoref{example-derivation}.

\begin{flalign*}
& \func{recv} \; : \; \Var_n \to \Process_n & \\
& \func{recv} \; c \; = \recv{c}{} (\recv{\suc c} \PO) \\
& \\
& \func{send} \; : \; \Var_n \to \Process_n & \\
& \func{send} \; c = \new{} \send{\suc c}{\constr{0}} \PO \\
\end{flalign*}
\begin{flalign*}
& \func{sync} \; : \; \Var_n && \to \Var_n \to \Var_n \to \Process_n & \\
& \func{sync} \; in0 \; in1 \; out \; &&= \\
& && \recv{in0}{} \\
& && \recv{(\suc \; in1)}{} \\
& && \send{(\suc \suc \; out)}{\suc \constr{0}} \\
& && \send{(\suc \suc \; out)}{\constr{0}} \PO \\
\end{flalign*}
\begin{flalign*}
& \func{example} && : \; \Process_{\constr{0}} & \\
& \func{example} && = \new{} ( \func{send} \; \constr{0} \\
& && \comp{}{ \new{} ( \func{send} \; \constr{0} } \\
& && \comp{}{ \new{} ( \func{recv} \; \constr{0} } \\
& && \comp{}{ \func{sync} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}))) } \\
\end{flalign*}

@@

\autoref{fig:algebra-instances} sketches the implementation of linear, graded and shared types as instances of our usage algebra.
Their use in typing derivations is illustrated in \autoref{example-derivation}.

@@

\begin{example}[Typing derivation]
\label{example-derivation}

@@

\begin{flalign*}
& \func{\vdash sync} && : \types{\gamma \comma \channel{t}{\lz} \comma \channel{t}{\lz} \comma \channel{t}{\lz}\\ & &&}{\Gamma \comma (1 \comma 0) \comma (1 \comma 0) \comma (0 \comma 2)\\ & &&}{\func{sync} \; (\suc \suc \constr{0}) \; (\suc \constr{0}) \; \constr{0}\\ & &&}{\Gamma \comma (0 \comma 0) \comma (0 \comma 0) \comma (0 \comma 0)} & \\
& \func{\vdash sync} && = \\
& && \recv{\suc \suc \constr{0}}{} \\
& && \recv{\suc \suc \constr{0}}{} \\
& && \send{\suc \suc \constr{0}}{\suc \constr{0}} \\
& && \send{\suc \suc \constr{0}}{\constr{0}} \PO \\
\end{flalign*}

@@

\begin{flalign*}
& \func{\vdash recv} \; \types{\gamma \comma \channel{t}{\lz}}{\Gamma \comma (2 , 2)}{\func{recv} \; \constr{0}}{\Gamma \comma (0 \comma 2)} & \\
& \func{\vdash recv} \; = \; \recv{\constr{0}}{} \recv{(\suc \constr{0})}{} \PO \\
& \\
& \func{\vdash send} \; \types{\gamma \comma \channel{t}{\lz} \comma t}{\Gamma \comma (1 , 1) \comma \lz}{\func{send} \; (\suc \constr{0}) \; \constr{0}}{\Gamma \comma (1 \comma 0)\comma \lz} & \\
& \func{\vdash send} \; = \; \send{(\suc \constr{0})}{\constr{0}} \PO
\end{flalign*}

@@

\begin{flalign*}
& \func{\vdash example} && : \; \types{\constr{[]}}{\constr{[]}}{\func{example}}{\constr{[]}} & \\
& \func{\vdash example} && = \new{} \; \_ \; \_ \; \_ \; (( \new{} \; \_ \; \_ \; \_ \; \func{\vdash send}) \\
& && \comp{}{ \new{} \; \_ \; \_ \; \_ } \; (( \new{} \; \_ \; \_ \; \_ \; \func{\vdash send}) \\
& && \comp{}{ \new{} \; \channel{\unit}{\func{\omega}} \; \_ \; \_ } \; ( \func{\vdash recv} \\
& && \comp{}{ \func{\vdash sync} }))) \\
\end{flalign*}
